Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)

### 2021-10-18 analysis/manuscript.Rmd
thar she blows

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  pdf_document:
    includes:
      in_header: "preamble.tex"
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---
\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

<!-- @Onyango2019 -->

@@

<!-- @Miles2009 -->

@@

<!-- @Lambert2005 -->

@@

<!-- @Boudewyn2007 -->

<!-- @Peltola2011 -->

@@

**Do we talk about the weaknesses of BRIMS and how we used 
this to improve?**

```{r BRIMS, echo=FALSE, fig.cap="BRIMS residue", out.width = '100%'}
knitr::include_graphics("../img/BRIMS.png")
```

@@

@@

#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
hourly_truck_rate <- 175 #$/hr
solid_volume_factor <- 0.4 #from fortsv5
adj_trailer_capacity <- trailer_capacity * solid_volume_factor
specific_gravity <- 0.38 #lodgepole pine
odt_trailer <- specific_gravity * adj_trailer_capacity
truck_rate <- hourly_truck_rate / odt_trailer

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ (DM)
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

@@

#get prop of dec/con by twp
standing_time <- 20

Canada's National Forest Inventory System (NFIS) is a product of the Canadian
Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
identifies locations of forest harvesting across Canada for the years 1985-2015
detected from Landsat satellite imagery.  These data are at a 30 m spatial
resolution.  The data were downloaded from
https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
(accessed 2020-03-05). The data set is described in Hermosilla *et al.* (2016)
@Hermosilla2016.
The data were clipped to the Alberta provincial boundary which was obtained from
Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of
the spatial information belonging to the Alberta government.

## Alberta township coverage
For our analysis, most of our information was aggregated to the township level.
In Alberta, the legal description of any parcel of land uses the Alberta
Township Survey System (see
https://www.alberta.ca/alberta-township-survey-system.aspx, accessed
2021-10-08). Most townships are roughly square in shape, with an average area of
about 9,533 ha.  The GIS coverage for township outlines was obtained from
Altalis (https://www.altalis.com, accessed 2021-09-10).
The harvest areas obtained from the NFIS forest harvesting data set were
aggregated to the township level so that the harvest area for each township in
each of the years from 1985 to 2015 was recorded in a data set.

## Hypothetical plant locations and capacities
There are a number of companies with Forest Management Agreement Areas
located near Edson, Whitecourt, and Hinton.
The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser
Mills Ltd (Edson) FMA is the primary timbershed for the Edson Forest Products 
sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMA supplies 
the Edson oriented strand board mill and the Drayton Valley sawmill.  The 
Millar Western FMA supplies sawmills in Whitecourt and Fox Creek, a pulp mill
in Whitecourt, and a specialty wood products facility and Acheson.  The Blue
Ridge Lumber Inc. FMA supplies a sawmill in Blue Ridge, which is about 23 km 
from Whitecourt.  The ANC Timber Ltd. FMA supplies logs to several sawmills in
the area, in exchange for chips used to feed the Alberta Newsprint
Company newsprint mill in Whitecourt.

```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'}
knitr::include_graphics("../img/west_central_AB_2.png")
```

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 000 Mg/a (DM) corresponding to a small wood pellet
mill, 400 000 Mg/a (DM) corresponding to a large pellet mill or a small ethanol
plant, and 800 000 Mg/a (DM) corresponding to a large ethanol plant.

```{r bb, echo=FALSE, fig.cap="Boreal Bioenergy--Obed.", out.width = '80%'}
knitr::include_graphics("../img/BorealBioenergy.png")
```

## Administrative and ecological boundaries.

### FMU map

@@

```{r FMU, echo=FALSE, fig.cap="Location of Forest Management Units in Alberta showing Green and White Areas.", out.width = '80%'}
knitr::include_graphics("../img/FMU.png")
```

@@

@@

```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
knitr::include_graphics("../img/ecozones.png")
```
Example output from the biomass calculator.

```{r biomass_calculator_example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
knitr::include_graphics("../img/biomass_calc_example.png")
```

```{r taper, echo=FALSE, fig.cap="Volume calcualation example.", out.width = '80%'}
knitr::include_graphics("../img/taper.png")
```
## Spatial variability in residual biomass production

@@

## Service areas

The provincial road network was sorted into four classes: primary highways,
other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
Loaded and unloaded travel speeds were assumed for each road class, according
to the table below. The harmonic mean of loaded and unloaded speeds 
(rounded to the nearest
multiple of 5 km/h) were calculated as

$s_h= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}$.

Use of harmonic mean allows us to calculate the correct round trip travel time.

Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to
try to format this table better.

* Rasterized network areas were generated using the QGIS plugin, QNEAT3
(Iso-Area as Interpolation) @Raffler2018. **How big are the cells?**

* QGIS zonal statistics were used to calculate the median network travel time in
each township.

```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time from Edson, limited to 6 hours.", out.width = '100%'}
knitr::include_graphics("../img/network_service.png")
```

@@

## Residue yield (Mg · ha^-1^ (DM)) by township.

@@

\[
d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
\]

@@

\[
H(d_b) = \beta_0 d_b ^ {\beta_1}
\]

@@

\[
a(d_b,h) = \pi \left(\frac{d(d_b,h)}{200}\right)^2
\]

\[
v(d_b,h_s,h_t) = \int_{h_s}^{h_t}\pi \left(\frac{d(d_b,h)}{200}\right)^2dh
\]

The constant 200 is used to convert a diameter (cm) to a radius (m).  
**find a way to reduce possible confusion with $d$**

@@

### Township level biomass

For both needleleaf and broadleaf, we multiply the residue associated with the
individual tree by the number of harvested stems per ha.  We reduce this
value using a recovery factor of 0.645 **??** following Peltola *et al.* 
@Peltola2011, recognizing that some of the residual biomass will 
be unrecoverable.

## Cost calculations

```{r roser, echo=FALSE, fig.cap="Röser.", out.width = '100%'}
knitr::include_graphics("../img/roser.png")
```

```{r hourly_rates, echo=FALSE, fig.cap="Hourly rates for forestry equipment.", out.width = '100%'}
knitr::include_graphics("../img/hourly_rates.png")
```

Following @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ (DM), the comminution (*grinding ?*) cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ (DM), and other costs were
set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ (DM).  The total "constant" costs are
`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ (DM).
Using *blah blah blah* , the truck rate was set to 
`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^ (DM)/h.
A 100 m³ trailer can haul `r odt_trailer` Mg (DM) of chipped residue.

@@

```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
#knitr::include_graphics("../img/alberta_harvest.png")
harv_comparison
```

```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'}
knitr::include_graphics("../img/harvestbyowner.png")
```

Some lovely supply curves

```{r supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (cycle time) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}

p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg (DM))") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,12))

p

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg (DM))") +
  labs(y = "marginal cost ($·Mg^-1^ (DM))") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,175)) 
```

```{r boxplots, echo=FALSE, warning=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
t200$capacity <- 200
t400 <- cum.odmt %>% filter(residue.cumsum <= 400000)
t400$capacity <- 400
t800 <- cum.odmt %>% filter(residue.cumsum <= 800000)
t800$capacity <- 800

@@

p <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($·Mg^-1^ (DM))", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))
p

@@

```{r avecosts, echo=FALSE, warning=FALSE, fig.cap="Average costs by location, capacity, and year "}

@@

ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($·Mg^-1^ (DM))") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,150))

@@

# Results and discussion

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  word_document:
     number_sections: yes
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    includes:
      in_header: "preamble.tex"
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---
\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

@@

@@

@@

@@

@@

## Data sources

We used a number of publicly available datasets to develop the datasets
used in our analysis.  Most of the data are freely available from the World
Wide Web (WWW). Much of the data processing required for this study was 
manipulation of spatial
data.  This was done using the QGIS software @QGISorg2021.
All national level data were clipped to the Alberta provincial boundary
for further analysis.

The datasets are only introduced here.  The way they are used will be discussed
in detail later.

### Altalis

Altalis manages much of the spatial data created by or for the Government of
Alberta.  We use several coverages from their base features database product 
@Altalis2021:
the provincial, green/white area, forest management unit (FMU), forest management
agreement area (FMAA), Alberta Township Survey System
township, and town
boundaries.

The Alberta Township Survey System (ATS) is particularly important to this study
as all the data used for this study are aggregated or disaggregated to the
township level.  ATS provides the basis for legal land descriptions in Alberta.
The typical township is roughly square (9.778 km by 9.716 km, 9 500 ha) and 
is indexed by
township (1--126), range (1--30), and meridian (4--6) numbers.  
The townships immediately east of
Alberta's western boundary, and immediately east of the 5^th^ and 6^th^
meridians (corresponding to 114°W and 118°W longitude, respectively) are 
partial townships and therefore deviate from the roughly square shape.  ATS
is described in detail by the Alberta Land Surveyor's Association @ALSA2021.

### Alberta Geospatial Services

Alberta Geospatial Services is operated by the Alberta Ministry of Environment
and Parks @Alberta2021a.  We obtained the road access coverages from here.  The
road access coverages available from Altalis do not have as detailed information
on road class as those available from Alberta Geospatial Services.

### National Forest Information System

As part of Canada's National Forest Inventory, the National Forest Information
System (NFIS) provides a web-based individual tree biomass calculator @NFIS2021. 
It 
is used to calculate stem wood, stem bark, branch, and foliage biomass for 
a tree based on province, ecozone, species, diameter at breast height (dbh) and
height.  It is possible to upload a file representing many trees with different
combinations of these input parameters and download a file containing the
calculated results for all the trees.

NFIS has also produced a set of spatial data layers representing forest 
properties at a 250 m resolution for all of Canada @Beaudoin2017.  We used the
merchantable volume, stand age, and species composition attributes from these 
data layers.

Another product of NFIS is a set of spatial datasets derived from satellite data 
@NFIS2021b.  The dataset used for this study is the harvest year data set for
the years 1985-2015.  The dataset is described in detail by Hermosilla *et al.*
@Hermosilla2016.  Essentially, it consists of a raster for Canada at 30 m 
resolution indicating the year of harvest for each cell.

### Natural Resources Canada

For our study, the most important product of Natural Resources Canada is
the Ung *et al.* paper @Ung2013 which describes the tree taper models 
developed for Canada.  A valuable contribution related to this is the WWW-based
wood volume calculator that uses those models @NRCan2015.  It is helpful for
understanding how taper functions work.

### National Forestry Database

The national forestry database @CCFM2020 was used to summarize the 
timber harvest volume from
Alberta for the years 1990-2015.

### Forest management plans

The Alberta government requires FMA holders to develop a forest management
plan for their FMAA.
The forest management plans were used to get information on utilization 
standards and piece size of harvested trees @Alberta2021.

### Miscellaneous

The map of provincial boundaries comes from Statistics Canada @Canada2016.

The boundaries of Canada's terrestrial ecozones, which are important inputs to
the biomass calculator can be found on Agriculture and Agri-Food Canada's
web site @Canada2017.

## Study area

Alberta is a province in western Canada.  It occupies 661 848 km^2^.
Its southern border is 49°N latitude, northern border is 60°N latitude,
eastern border is 110°W longitude, and its western boundary is defined by
120°W longitude and the Great Divide of the Rocky Mountains.

```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '80%'}
knitr::include_graphics("../img/canada.png")
```

The southeastern part of the province is largely privately owned agricultural
land, but the northern and western parts are largely publicly owned forest land.
There is a substantial area of privately owned agricultural land
in the Peace River Country of northwestern Alberta.  
There are also substantial areas
in National Parks along the Rocky Mountains and in the far north of the 
province.  The province calls the area of mostly publicly owned forest land 
as the Green Area, and the area of mostly privately owned agricultural land 
as the White Area.  About 90% of the harvested timber volume in Alberta
comes from the Green Area.

For administrative purposes, the Green Area is divided into Forest Management
Units (FMUs).  The province also enters into long term agreements with forestry
companies giving the companies the rights to timber harvest and the 
responsibility to manage the forest sustainably.  These agreements are
called Forest Management Agreements, and the areas under agreement are
called Forest Management Agreement Areas (FMAAs).  For the most part, the
boundaries of the FMAAs are now coincident with FMU boundaries, with
the exception that the Alberta-Pacific FMAA in eastern Alberta comprises
12 FMUs.
<!-- FMAAs occupy 62% of the Green Area **and maybe something about proportion of -->
<!-- harvest** -->

Our focus for this study is the forest near Hinton, Edson, and Whitecourt in 
west-central Alberta.  These towns were chosen as they host several
forest products companies and are located near very active FMAAs.
We assume that it would be likely that a forest residue processing plant
would be located near an existing forest products mill as the forest road
network has been developed to supply those mills, and location of a residue
plant near a forest products mill would also allow for access to 
mill residues.  Figure ?? shows the location of the three towns in 
relation to FMAAs and the Green and White Areas.

The FMAAs which supply most of the timber to mills in our near our three 
locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills Ltd. (Edson),
Weyerhaeuser Company Limited (Pembina Timberlands), Millar Western, Blue
Ridge Lumber Inc., and ANC Timber Ltd.

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 000 Mg/a  corresponding to a small wood pellet
mill, 400 000 Mg/a  corresponding to a large pellet mill or a small ethanol
plant, and 800 000 Mg/a  corresponding to a large ethanol plant.

```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'}
knitr::include_graphics("../img/study_area.png")
```

The areas harvested in Alberta in 2009 according to NFIS @NFIS2021b
@Hermosilla2016 are shown in relationship to the three towns and 
the Green Area in Figure 3.  Our data set includes harvest areas for the
years 1990--2015, so this is just an example.

```{r cutblocks2009, echo=FALSE, fig.cap="Harvest areas for 2009.", out.width = '80%'}
knitr::include_graphics("../img/cutblocks2009.png")
```

Figure 4 shows the proportion of the township area harvested in 2009 aggregated
to the township level, showing only those townships where the proportion of
the township area harvested was greater than 0.001.  For each township, and
each of the years between 1990 and 2015, we store the area harvested (ha) for
use in our analysis.  We can see from this map that a great deal of timber
harvest took place near our plant locations in 2009.

```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.  Only those townships where the proportion of total area harvested was 0.001 or greater are shown", out.width = '80%'}
knitr::include_graphics("../img/harvesttwp.png")
```

## Cycle time estimates

The provincial road network was sorted into four classes: primary highways,
other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
Loaded and unloaded travel speeds were assumed for each road class, according
to the table below. The harmonic mean of loaded and unloaded speeds 
(rounded to the nearest
multiple of 5 km/h) were calculated as
\begin{equation}
\bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
\end{equation}
where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
Use of harmonic mean allows us to calculate the correct  cycle time:  the
time required to travel from the mill, to the collection site in the forest,
and back to the mill.  This provides the basis for our estimates of haul cost.
Calculating average haul speed using an arithmetic mean leads to a biased
estimate of cycle time.

<!-- Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to -->
<!-- try to format this table better. -->

Figure 6 illustrates the road network in Alberta and estimates of one-way haul
time based on the average of loaded and unloaded speeds from Table ??. The
road network in the White Area of the province is very dense and largely arranged
as a grid related to the Alberta Township Survey System.  Most of the roads
in the White Area are public.  The road network in the Green Area is much
less dense and many of the roads are built and maintained by resource industries 
(*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
generated using the QGIS plugin, QNEAT3
(Iso-Area as Interpolation) @Raffler2018.
QGIS zonal statistics were used to calculate the median network travel time to
each township, from each of our three towns.  The one-way travel time in Figure
6 was limited to 6 hours because this equates to a 13 hour cycle time if
you assume 1 hour for loading and unloading.  In Alberta, drivers of commercial
vehicles are not permitted to drive more than 13 hours in a day.  **technically
we could take the picture out to 6.5 hours, as the loading and unloading time
does not count as driving time.  There is an on-duty limit of 14 hours**.

```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
knitr::include_graphics("../img/network_service.png")
```

@@

#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
hourly_truck_rate <- 175.90 #$/hr
#solid_volume_factor <- 0.4 #from fortsv5
solid_volume_factor <- 0.3 #from fortsv5
adj_trailer_capacity <- trailer_capacity * solid_volume_factor
specific_gravity <- 0.38 #lodgepole pine
odt_trailer <- specific_gravity * adj_trailer_capacity
truck_rate <- hourly_truck_rate / odt_trailer

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

@@

#get prop of dec/con by twp
standing_time <- 60

```
### Utilization standards

@@

## Biomass collection costs

All measures of biomass (Mg) used in this study are reported on
a dry matter basis.

We assume that the biomass available for collection at a cutblock includes the
tops, branches, and foliage left at a roadside landing after processing by
a stroke delimber.  The delimber is assumed to top the log at the top diameter
specified in the utilization standard relevant to the harvest area.  Recovery
will not be perfect.  We assume a recovery factor of 0.624 
following Peltola *et al.* 
@Peltola2011, recognizing that some of the residual biomass will 
be unrecoverable.

Our focus in this study is on the spatial and temporal variation in biomass
collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
(roads, supervision, overhead, maintenance, and compliance) were
set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .  The total cost,
excluding transport, is
`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ .

We assume that the residue will be transported from the forest using a 
live floor chip van with 100 m³ capacity.
The USDA Forest Service's Forest Residue Transportation Costing Model 
@Rummer2005 assumes a default solid volume factor of 0.3 for biomass meaning that
a 100 m³ capacity trailer could hold the equivalent of 30 m³ of solid biomass.

This equates to `r sprintf("%5.2f",odt_trailer)` Mg
of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
according to Miles and Smith @Miles2009a.

Timber Tracks produces a publication describing forestry equipment hourly rates
@TT2020.  The rate reported for an 8 axle tridem tractor with a
b-train trailer was `r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
the tractor-trailer combination used to haul biomass would cost a 
similar amount.

Therefore, the cost of transporting residual biomass was set to 
`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^·h^-1^.

<!-- ## Hypothetical plant locations and capacities -->
<!-- Location stuff based on Onyango @Onyango2019. -->

<!-- There are a number of companies with Forest Management Agreement Areas -->
<!-- located near Edson, Whitecourt, and Hinton. -->
<!-- The West Fraser Mills Ltd. (Hinton) FMAA is the primary timbershed for the -->
<!-- Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the Edson Forest Products  -->
<!-- sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies  -->
<!-- the Edson oriented strand board mill and the Drayton Valley sawmill.  The  -->
<!-- Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies sawmills in Whitecourt and Fox Creek, a pulp mill -->
<!-- in Whitecourt, and a specialty wood products facility and Acheson.  The Blue -->
<!-- Ridge Lumber Inc. FMAA supplies a sawmill in Blue Ridge, which is about 23 km  -->
<!-- from Whitecourt.  The ANC Timber Ltd. FMAA supplies logs to several sawmills in -->
<!-- the area, in exchange for chips used to feed the Alberta Newsprint -->
<!-- Company newsprint mill in Whitecourt. -->

<!-- ```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'} -->
<!-- knitr::include_graphics("../img/west_central_AB_2.png") -->
<!-- ``` -->

@@

Most
of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs
in the Montane Cordillera and Taiga Plain.

@@

```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
knitr::include_graphics("../img/ecozonesAlex.png")
```

## Spatial variability in residual biomass production

@@

@@

## Residue yield (Mg · ha^-1^ ) by township.

@@

\begin{equation}
d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
\end{equation}

@@

\begin{equation}
H(d_b) = \beta_0 d_b ^ {\beta_1}
\end{equation}

@@

\begin{equation}
r(d_b,h) = \frac{d(d_b,h)}{200}
\end{equation}

The constant 200 is used to convert a diameter (cm) to a radius (m).

\begin{equation}
a(d_b,h) = \pi r(d_b,h)^2
\end{equation}

\begin{equation}
v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
\end{equation}

@@

@@

```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
suppressWarnings(print(harv_comparison))
```

<!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
<!-- knitr::include_graphics("../img/harvestbyowner.png") -->
<!-- ``` -->

# Results and Discussion

## Township level biomass

## Distributional results
```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}

cyclePlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg )") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,14))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
mcPlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg )") +
  labs(y = "marginal cost ($·Mg^-1^ )") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,250)) 
```

```{r boxplots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
t200$capacity <- 200
t400 <- cum.odmt %>% filter(residue.cumsum <= 400000)
t400$capacity <- 400
t800 <- cum.odmt %>% filter(residue.cumsum <= 800000)
t800$capacity <- 800

@@

bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($·Mg^-1^ )", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))

@@

```{r avecosts, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Average costs by location, capacity, and year "}

@@

avePlot <- ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($·Mg^-1^ )") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,200))

@@

```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location.", out.width = '100%'}
suppressWarnings(print(cyclePlot))
```

```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location.", out.width = '100%'}
suppressWarnings(print(mcPlot))
```
